Visualizing Conditional Relationships

### 2017-08-30 paper/main.tex
cleaned up bibliographies

\section{Introduction}
\label{sec:introduction}
\begin{figure}
  \includegraphics{chart_chooser.png}
  \caption{From Bertin's Semiology of Graphics \cite{bertin_semiology_2011}, this diagram
    illustrates how there are many ways to display the same information about the 1954 Frnech
  workforce.}
  \label{fig:chart_chooser}
\end{figure}

@@

While this survey focuses on data structured in a linear, tabular, or
data cube way\cite{munzner_what:_2014}, it could potentially be expanded to network data.

%%\subfile{sections/distdist.tex}
%%\subfile{sections/distdens.tex}
%%\subfile{sections/densdist.tex}
%%\subfile{sections/densdens.tex}

>>>>

\section{Introduction}
\label{sec:introduction}
\begin{figure}
  \includegraphics[width=\textwidth]{chart_chooser.png}
  \caption{From Bertin's Semiology of Graphics \cite{bertin_semiology_2011}, this diagram
    illustrates how there are many ways to display the same information about the 1954 French
  workforce.}
  \label{fig:chart_chooser}
\end{figure}

@@

While this survey focuses on data structured in a linear, tabular, or
data cube way \cite{munzner_what:_2014}, it could potentially be expanded to network data.

\subfile{sections/distdist.tex}
%%\subfile{sections/distdens.tex}
%%\subfile{sections/densdist.tex}
%%\subfile{sections/densdens.tex}

### 2017-08-30 paper/sections/densdens.tex
cleaned up bibliographies

\begin{figure}
  \includegraphics{partial_dependence}
  \caption{Taken from Scikit Learn\cite{scikit_learn}, the line plots show how California housing prices are partially dependent on each x variable. The heat map illustrates the joint effect of the age the house and the number of occupants in the house.}
  \label{fig:partialdependence}
\end{figure}

One of the most straightforward ways to show dependencies is to plot one
variable against another, but those methods often exclude the effects of the
other variables in the datasets. Regression methods can be used to analyze the
ways in which one variable is partially dependent on many other variables in
the dataset\cite{Hastie_Tibshirani}. As shown in figure~\ref{fig:partialdependence}, one way partial dependence line plots and two way partial dependence heat maps show how California housing prices are marginally dependent on each variable.

While bar graphs and histograms can give a sense of the variability in
data, they are limited to showing the dispersion of the data only in how the measurements are spread out. This is somewhat limiting for very large datasets, so the box plot has evolved as a way to display more characteristics of the distribution \cite{wickham2011}. While a simple box plot does not inherently show conditional dependencies, many derivations do.

\begin{figure}
  \includegraphics{boxplot}
  \caption{Tukey's 1977 example of the box plot (exhibit 6 of chapter 2 from
    Exploratory Data Analytics\cite{tukey1977}) illustrates how the box plot
    shows
    the distribution of topology of states and
    the distribution of volcanoes. The strength of the box and whisker is that
    the
    reader can quickly compare the two distributions and learn, for example,
    that
    volcanoes tend to span a shorter range of average heights, but that the
    extremes
    are much further from the center. Essentially, the distributions act as
    references for the other displayed distributions.}
  \label{fig:boxplot}
\end{figure}

Tukey's introduction to the box-plot \cite{tukey1977}, shown in
Figure~\ref{fig:boxplot} illustrates how the distributions of these 1 dimensional densities differ by category. While this graph shows the median, range between upper and lower interquartile, and lower and upper extreme, many authors have exploited the structure to convey more information about the data. Authors used different quantile levels \cite{hyndman1996}s or measures of outliers
\cite{frigge1989}, or ways of identifying outliers\cite{carter2009,
  schwertman2004, schwertman\
  2007}, or otherwise incorporated skewness,
kurtosis, and other descriptive distributional statistics \cite{kim2004,
  hubert2008, marmolejo\
  2015}.

While box plots were designed to visualize discrete observations, often there
are no discrete observations and instead the `observation` is a timeseries
curve (where the connections are as important as the individual points) or
spatial region. Understanding the distributions of functional observations
often relies first ordering the functions such that there's some measure of how
close these functions are to each other--functional depth\cite{febrero2007},
bivariate score depth\cite{hyndman2009}, and
bivariate kernel density estimation\cite{scott1992} are some examples. This
ordering then yields functional analogs to median and other percentile
metrics. For small values, the ordered curves can then be colored based on the
ordering metric \cite{hyndman2009}, but this method is unwieldy for larger
ensembles.

\begin{figure}
  \includegraphics[width=.75\linewidth]{contour_weather}
  \caption{From \cite{whitaker2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the temporal dimension of the data.}
  \label{fig:countour}
\end{figure}
Functional bag, box, and HDR plots \cite{hyndman2009, sun2011}} tend to be good at visualizing specific types of outliers, but are somewhat limited to only one functional distribution. A very common task is to compare ensembles, each a function of the same timeseries or geographic region. Contour box plots\cite{whitaker2013} aim to capture the variability of the
uncertainty via a band depth ordering metric. Each ensemble members band depth
is computed as sum of the probabilities that the observations in any given
ensemble fall within the max-min envelope defined by any two other
ensembles. The bands are then sorted by band depth such that the median is the
ensemble with about 50\% of its members whiten all envelopes formed by other
bands (so most centered). Outer bands are chosen according to the task at
hand, which in this paper is temperature visualizations. The authors argue that contour box plots are an improvement over spaghetti plots \cite{lau_koa_pang} because the spaghetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns. Instead in
Figure~\ref{fig:countour}, the authors remove most of the bands and instead visualize an outlying envelope(light gray) and a more central envelope(dark gray). It retains much of the information of
the spaghetti, but removes the visual noise of the lines.

One of the most common approaches for visualizing datasets that are structurally complex or have more than one or two levels of conditional dependencies is to built a toolkit comprising of many different basic visualization types. The calendar visualization described above is a simple example, and just like more complex tools it follows the design principal of overview-zoom/filter-details-on-demand \cite{shneiderman1996eyes} because these tools facilitate engagement with the data at the aggregate and encourage drilling down to the core. 
\end{document}

>>>>

\begin{figure}
  \includegraphics{partial_dependence}
  \caption{Taken from Scikit Learn\cite{_partial_????}, the line plots show how California housing prices are partially dependent on each x variable. The heat map illustrates the joint effect of the age the house and the number of occupants in the house.}
  \label{fig:partialdependence}
\end{figure}

One of the most straightforward ways to show dependencies is to plot one
variable against another, but those methods often exclude the effects of the
other variables in the datasets. Regression methods can be used to analyze the
ways in which one variable is partially dependent on many other variables in
the dataset\cite{_elements_2009}. As shown in figure~\ref{fig:partialdependence}, one way partial dependence line plots and two way partial dependence heat maps show how California housing prices are marginally dependent on each variable.

While bar graphs and histograms can give a sense of the variability in
data, they are limited to showing the dispersion of the data only in how the measurements are spread out. This is somewhat limiting for very large datasets, so the box plot has evolved as a way to display more characteristics of the distribution \cite{wickham_40_2011}. While a simple box plot does not inherently show conditional dependencies, many derivations do.

\begin{figure}
  \includegraphics{boxplot}
  \caption{Tukey's 1977 example of the box plot from
    Exploratory Data Analytics\cite{tukey_exploratory_1977} illustrates how the box plot shows the distribution of topology of states and
    the distribution of volcanoes. The strength of the box and whisker is that the
    reader can quickly compare the two distributions and learn, for example,
    that volcanoes tend to span a shorter range of average heights, but that the
    extremes are much further from the center. Essentially, the distributions act as references for the other displayed distributions.}
  \label{fig:boxplot}
\end{figure}

Tukey's introduction to the box-plot \cite{tukey_exploratory_1977}, shown in
Figure~\ref{fig:boxplot} illustrates how the distributions of these 1 dimensional densities differ by category. While this graph shows the median, range between upper and lower interquartile, and lower and upper extreme, many authors have exploited the structure to convey more information about the data. Authors used different quantile levels \cite{hyndman_sample_1996}s or measures of outliers
\cite{frigge_implementations_1989, schwertman_identifying_2007}, or otherwise incorporated skewness, kurtosis, and other descriptive distributional statistics \cite{kim_more_2004, marmolejo-ramos_shifting_2015}.

While box plots were designed to visualize discrete observations, often there
are no discrete observations and instead the `observation` is a timeseries
curve (where the connections are as important as the individual points) or
spatial region. Understanding the distributions of functional observations
often relies first ordering the functions such that there's some measure of how close these functions are to each other--functional depth\cite{febrero_functional_2007},
bivariate score depth\cite{rob_j._hyndman_rainbow_2010}, and
bivariate kernel density estimation\cite{scott_multivariate_1992} are some examples. This
ordering then yields functional analogs to median and other percentile
metrics. For small values, the ordered curves can then be colored based on the
ordering metric \cite{rob_j._hyndman_rainbow_2010}, but this method is unwieldy for larger
ensembles.

\begin{figure}
  \includegraphics[width=.75\linewidth]{contour_weather}
  \caption{From \cite{whitaker_contour_2013}, the contour box plot method is used to show uncertainty in temperature ensembles while maintaining the temporal dimension of the data.}
  \label{fig:countour}
\end{figure}
Functional bag, box, and HDR plots \cite{rob_j._hyndman_rainbow_2010, sun_functional_2011} tend to be good at visualizing specific types of outliers, but are somewhat limited to only one functional distribution. A very common task is to compare ensembles, each a function of the same timeseries or geographic region. Contour box plots\cite{whitaker_contour_2013} aim to capture the variability of the
uncertainty via a band depth ordering metric. Each ensemble members band depth
is computed as sum of the probabilities that the observations in any given
ensemble fall within the max-min envelope defined by any two other
ensembles. The bands are then sorted by band depth such that the median is the
ensemble with about 50\% of its members whiten all envelopes formed by other
bands (so most centered). Outer bands are chosen according to the task at
hand, which in this paper is temperature visualizations. The authors argue that contour box plots are an improvement over spaghetti plots \cite{luo_visualizing_2003} because the spaghetti plots get nosy as the number of plots increase and so it's hard to tease out specific patterns. Instead in
Figure~\ref{fig:countour}, the authors remove most of the bands and instead visualize an outlying envelope(light gray) and a more central envelope(dark gray). It retains much of the information of
the spaghetti, but removes the visual noise of the lines.

One of the most common approaches for visualizing datasets that are structurally complex or have more than one or two levels of conditional dependencies is to built a toolkit comprising of many different basic visualization types. The calendar visualization described above is a simple example, and just like more complex tools it follows the design principal of overview-zoom/filter-details-on-demand \cite{shneiderman_eyes_1996} because these tools facilitate engagement with the data at the aggregate and encourage drilling down to the core. 
\end{document}

### 2017-08-30 paper/sections/densdist.tex
cleaned up bibliographies

\begin{figure}
  \includegraphics{flexaxis.png}
  \caption{Figure taken from \cite{claessen_van_wijk_2011} illustrates the following common
    methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and radar charts.
  \label{fig:flexlink}
\end{figure}

There are a number of methods for showing how measurements of different
variables interact, and as showing figure~\ref{fig:flexlink} color is often
used to illustrate how these relationships are different for each category of
data. As seen in figure~\ref{fig:flexlink}, the common methods for showing
visualizations are matrix of scatters \cite{elmqvist, } , parallel
coordinates \cite{islenberg, wegman} and radar charts\cite{chambers_tukey}/

@@

Another important factor in visualizing categorical data is retaining the
inherent hierarchy in many datasets with categorical measurements
\cite{shneiderman_visualizing_2000}. The Parallel Sets tool
showin in figure~\ref{fig:parallelsets} is explicitly designed to show frequencies
of set membership in heirarichal ordered categorical sets, so the \{d_0, \dots, d_n} \in D
  \cite{kosara_parallel_2006}. Building on the flexible linked axis version of
  the parallel coordinates plot \cite{claessen_flexible_2011}, Parallel Sets
  treats each categorical set independently but indicates conditional
  dependency through cross tabulations and by grouping the data, indicated via
  color, based on an active dimension that then becomes the pivot point for the
  cross tabulations. In Parallel Sets, quantitative data is converted to
  categorical via binning. Many-to-many PCP is another type of generalized
  parallel coordinates where all possible pairs are explored \cite{lind}.

>>>>

\begin{figure}
  \includegraphics{flexaxis.png}
  \caption{Figure taken from \cite{claessen_flexible_2011} illustrates the following common
    methods for visualizing multivariate datasets: matrix of scatters, parallel coordinates and radar charts.
  \label{fig:flexlink}
\end{figure}

There are a number of methods for showing how measurements of different
variables interact, and as showing figure~\ref{fig:flexlink} color is often
used to illustrate how these relationships are different for each category of
data. As seen in figure~\ref{fig:flexlink}, the common methods for showing
visualizations are matrix of scatters \cite{elmqvist_rolling_2008} , parallel
coordinates \cite{inselberg_plane_1985, wegman_hyperdimensional_1990} and radar charts\cite{chambers_graphical_1983}.

@@

Another important factor in visualizing categorical data is retaining the
inherent hierarchy in many datasets with categorical measurements
\cite{shneiderman_visualizing_2000}. The Parallel Sets tool
showin in figure~\ref{fig:parallelsets} is explicitly designed to show frequencies
of set membership in heirarichal ordered categorical sets \cite{kosara_parallel_2006}. Building on the flexible linked axis version of
  the parallel coordinates plot \cite{claessen_flexible_2011}, Parallel Sets
  treats each categorical set independently but indicates conditional
  dependency through cross tabulations and by grouping the data, indicated via
  color, based on an active dimension that then becomes the pivot point for the
  cross tabulations. In Parallel Sets, quantitative data is converted to
  categorical via binning. Many-to-many PCP is another type of generalized
  parallel coordinates where all possible pairs are explored \cite{lind_many--many_2009}.

### 2017-08-30 paper/sections/distdens.tex
cleaned up bibliographies

\begin{figure}
\includegraphics{radial}
\caption{Figure from \cite{Albo_Bak_}. Radar chart, flower chart, and circle
  charts}
\label{fig:radial}
\end{figure}
Radial visualizations can be used to show representative
categories from multivariate vectors. For example, as seen in figure~\ref{fig:radial}
radar plot, the clusterings of the area types indicates general categories for
the data, and the number of areas in each cluster indicates the relative
frequency. Flower charts can also be grouped to find these relative
categories, as can the rings in a circle chart. Parallel coordinates charts are often explicitly grouped by category, but the relative frequency of the colors can be used to derive the relative frequencies of each graph.

\begin{figure}
\includegraphics{calendar.png}
\caption{Figure from \cite{van_wijk_van_selow}. They visualized representative clusters of electricty usage and the relative frequency of each cluster using a calendar where day and cluster that day share the same color.}
\label{fig:calendar}
\end{figure}

Finding patterns and trends in large datasets, whether univariate or multivariate, is often a matter of trying to find representational fingerprints of common or outlying events in the dataset. This is frequently done through the use of a clustering algorithm. Figure~ref{fig:calendar} shows an analysis of electricity usage in the Netherlands\cite{van_wijk_van_selow}. Because the individual observations were recorded every 6 hours for a year, van Wilk and van Selow transformed the hourly observations into a matrix where each vector holds the observations for a given day. They used an agglomerative clustering algorithm \cite{kaufman_rousseeuw} to find representative days, and color matched these clusters to a calendar of days to visualize how these clusters were distributed over time.

>>>>

\begin{figure}
\includegraphics{radial}
\caption{Figure from \cite{albo_off_2016}. Radar chart, flower chart, and circle
  charts}
\label{fig:radial}
\end{figure}
Radial visualizations can be used to show representative
categories from multivariate vectors. For example, as seen in figure~\ref{fig:radial}
radar plot, the clusterings of the area types indicates general categories for
the data, and the number of areas in each cluster indicates the relative
frequency. Flower charts can also be grouped to find these relative
categories, as can the rings in a circle chart. Parallel coordinates charts are often explicitly grouped by category, but the relative frequency of the colors can be used to derive the relative frequencies of each graph.

\begin{figure}
\includegraphics{calendar.png}
\caption{Figure from \cite{van_wijk_cluster_1999}. They visualized representative clusters of electricty usage and the relative frequency of each cluster using a calendar where day and cluster that day share the same color.}
\label{fig:calendar}
\end{figure}

Finding patterns and trends in large datasets, whether univariate or multivariate, is often a matter of trying to find representational fingerprints of common or outlying events in the dataset. This is frequently done through the use of a clustering algorithm. Figure~ref{fig:calendar} shows an analysis of electricity usage in the Netherlands\cite{van_wijk_cluster_1999}. Because the individual observations were recorded every 6 hours for a year, van Wilk and van Selow transformed the hourly observations into a matrix where each vector holds the observations for a given day. They used an agglomerative clustering algorithm \cite{kaufman_agglomerative_1990} to find representative days, and color matched these clusters to a calendar of days to visualize how these clusters were distributed over time.

Techniques specifically aimed at clustering functional data can be found in the survey by Jacques, Julian and Preda \cite{Jacques_functional_2014}. %%todo: read and write up

### 2017-08-30 paper/sections/distdist.tex
cleaned up bibliographies

\documentclass[../main.text]{subfiles}
\begin{document}

@@

Categorical data can be likened to set measurement, where each category is a set
and researchers often want to know how many members each set has and what
intersections are common between sets \cite{agresti_categorical_2011,schneider_set-theoretic_2012}
  It is the latter that is analogous to a
conditional probability, the question of the likelihood of membership in one
set given membership in another. And sometimes there is also a quantitative
component involved too. The most common approach is usually to create a list of frequencies, sometimes of sets of one category, such as \D, or a cross-table\cite{goodman} where each cell is a an element in the product of two (or more sets), such as \D\cross \I creates a table
 where cell one is the frequency of elements that are in both $\d_0$ and
 $\i_0$.

 While a table is encouraged for a small number of variables
 \cite{munznertablesgood}), bar graphs and histograms are typically used to display
 differences between subsets such as \d_0,\dots, \d_n
 \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms
 include the hanging rootgram, ord plot, and other distribution oriented
 plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}. While conditional  distributions can somewhat be visualized using Fourfold displays, Mosaic displays, and using logit regression models, these plots have somewhat limited dimensionality and are somewhat limited to 3 sets of categories
 \cite{friendly_visualizing_2000}.

 \begin{figure}
   \includegraphics{upsetfig1}
   \caption{UpSet dashboard displaying the relationships amongst movie
     genres. On the left is the set visualization; the columns are the sets and
     the rows are the exclusive intersections or aggregates. On the right is
     the element view, and the scatter plot is a comparison of two sets with
     each dot equaling one element in the set.}
   \label{fig:upsetfig}
 \end{figure}

>>>>

\documentclass[../main.tex]{subfiles}
\begin{document}

@@

Categorical data can be likened to set measurement, where each category is a set
and researchers often want to know how many members each set has and what
intersections are common between sets \cite{agresti_categorical_2011,schneider_set-theoretic_2012}
  It is the latter that is analogous to a
conditional probability, the question of the likelihood of membership in one
set given membership in another. And sometimes there is also a quantitative
component involved too. The most common approach is usually to create a list of frequencies, sometimes of sets of one category or a cross-table\cite{goodman_measures_1991} which is an aggregate function applied to the product of two or more sets.

 While a table is encouraged for a small number of variables
 \cite{munzner_what:_2014}), bar graphs and histograms are typically used to display differences between subsets \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms
 include the hanging rootgram, ord plot, and other distribution oriented
 plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}. While conditional  distributions can somewhat be visualized using Fourfold displays, Mosaic displays, and using logit regression models, these plots have somewhat limited dimensionality and are somewhat limited to 3 sets of categories
 \cite{friendly_visualizing_2000}.

 %%\begin{figure}
  %% \includegraphics[width=\textwidth]{upsetfig1}
   %%\caption{UpSet dashboard displaying the relationships amongst movie
  %%genres. On the left is the set visualization; the columns are the sets and
    %% the rows are the exclusive intersections or aggregates. On the right is
    %% the element view, and the scatter plot is a comparison of two sets with
     %%each dot equaling one element in the set.}
   %%\label{fig:upsetfig}
 %%\end{figure}
